XSGRB - The X-shooter GRB afterglow sample

### 2017-01-06 document/XSGRB_sample.tex
BetaOX

\title{XSGRB - The X-shooter GRB afterglow sample}
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}

@@

From March 2005 to March 2016 there have been about 350 {\it Swift} bursts
complying with our sample selection criteria (see ``Immediate objective''), and
about half of them have measured redshifts. Among the latter subset, the team
proposing these observations has measured about two thirds of the redshifts,
mainly with FORS1/2 and X-shooter (see Fynbo et al.\ 2009, ApJS, 185, 526; 
Jakobsson et al.\ 2012, ApJ, 752, 62; Kr\"uhler et al.\ 2012, ApJ, 758, 46). Our
current aim is to build a sample superior to our previous low-resolution survey
(Fynbo et al.\ 2009, ApJS, 185, 526), both in terms of quantity and  quality of
the spectra. Our program started as guaranteed time observations during periods
84-91, and we have continued in open time since then.

@@

\begin{figure*}[!ht]
	\centerline{\includegraphics{figures/skymap.pdf}}
	\caption{Mollweide projection in galactic coordinates of the full sky showing
		the positions on the sky of the bursts presented in this work. The empty
		circles are the positions of all the \textit{Swift} bursts detected until the
		end of 2016. In blue is shown the position of the bursts fulfilling the sample
		criteria specified in Sect. \ref{samplecrit}. The red circles are added around
		GRBs which have been followed up by X-shooter since the commissioning of the
		instrument. Blue points with red circles represent GRBs that enter our sample
		and have X-shooter spectroscopy. The different samples are compared in Sect.
		\ref{results}. The background shows the dust maps presented in
		\citet{Schlegel1998} where the sample criteria cut with $A_V \lesssim 0.5$ mag
		is removed.}
	\label{fig:skymap}
\end{figure*}

@@

\begin{figure}
	\centerline{\includegraphics[width=8cm]{figures/resolution_paper.eps}}
	\caption{Green datapoints show the FWHM (km/s) of Gaussian fits to unresolved telluric absorption lines in the VIS spectra, as a function of the FWHM of a Gaussian fit onto the trace in spatial direction at  792 nm. The lower horizontal axis is in units pixels, the top axis in arc seconds. The red datapoints show a subsample of NIR spectra.
		The grey line shows a linear fit to the VIS datapoints. }
	\label{fig:res}
\end{figure}

@@

Using the observational characteristics of the 1244 \todo{Get number} bursts
observed by \textit{Swift} to date \cite{Gehrels2009}, and the X-ray derived
hydrogen column densities \citep{Evans2009}, we can quantify the degree to which
our sample is biased against the overall \textit{Swift} sample in terms of
intrinsic GRB properties. The values are queried from the online
\textit{Swift} database maintained\footnote{http://swift.gsfc.nasa.gov/archive/grb\_table/}.
Three main samples are of interest to compare to assess the completeness of the
follow-up campaign; the full \textit{Swift} sample consisting of all the bursts
observed by \textit{Swift} (1), the burst that fulfill the selection criteria
imposed in Sect. \ref{samplecrit} (2), and the bursts actually followed up with
X-shooter (3).

@@

GRB follow-up requires the immediate availability of the telescope in order to
successfully carry out spectroscopic observations. Out of the 158 \todo{Get
exact number} bursts in the statistical sample, 32($\sim$ 20 percent) \todo{Get
exact number} have not been observed due to conditions, not relating the GRB or
afterglow properties. The reasons includes unavailability of the telescope due
to technical maintenance such af mirror re-coating, a visitor rejecting the TOO
trigger, or bad weather at the observing site. Because this cut is irrespective
of the GRB properties, it is not expected that this should change sample
properties. Addressing whether this cut introduces a bias in the followed sample
is not trivial, but in terms of intrinsic GRB properties, the sample
distributions does not change by much, as shown in Sect. \ref{completeness}.
\todo{Should this be quantified somehow?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The nature of triggered GRBs without detected optical afterglow.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A subset of all GRBs exhibit no detectable or very faint optical afterglow \citep{Groot1998,
	Djorgovski2001, Fynbo2001}, which had been parametrized in term of "darkness"
\citep{Jakobsson2004, VanderHorst2009}. The X-ray properties of this subset has
previously been investigated \citep{DePasquale2003, Melandri2012}, finding a
slightly higher X-ray luminosity compared to the optically brights bursts
combined with similar prompt characteristics. This indicates along with
investigations of host galaxy extinction \citep{Greiner2011, Kruhler2011, Hjorth2012}, that the
extinction of the optical afterglow is primarily driven by the presence of dust,
local to the burst \todo{Local dust reference}.

Because we only follow up targets with either a clear afterglow or counterpart, correctly associating a galaxy with a burst when there is no optical afterglow is more difficult.

This should include observations where the acquisition image has been taken, but
observations were discontinued due to non-detection. What are the X-ray
properties of these bursts?

@@

\begin{figure*}
	\centering
	\includegraphics[page=1, width=16cm]{figures/HI_measurements.pdf}
	\caption{Measurements of the hydrogen column-densities for all bursts with a
	clear Lyman alpha absorption system. In solid black is shown the spectrum with
	black dotted giving the corresponding 1-$\sigma$ error. Black dashed shows zero
	flux density. The solid red line is the absorption of column density equal to
	the value presented in Tab. \ref{HI_tab} with the 1-$\sigma$ interval shown
	with dashed lines.}
	\label{fig:HI1}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=2, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI2}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=3, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI3}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=4, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI4}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=5, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI5}
\end{figure*}

@@

\end{acknowledgements}

@@

\newpage
\appendix

@@

\subsection{GRB090313 (z = 3.373)}

\subsection{GRB090530 (z = 1.266)}

\subsection{GRB090809 (z = 2.737)}

\subsection{GRB090926 (z = 2.106)}

\subsection{GRB091018 (z = 0.971)}

\subsection{GRB091127 (z = 0.49)}

@@

\subsection{GRB100219A (z = 4.667)}
The data presented here also formed the basis of GCN \#
10441\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10441.gcn3}} and is published
in \citet{Thone2013}. Observations started 12.5 hours after the \textit{Swift}
trigger and has a total exposure time of 4.8 ks. Absorption features, including
those of \lya, from a multitude of ions are detected against the afterglow
continuum at $z = 4.667$. Additionally, absorption from an intervening system
is found at $z = 2.181$.

>>>>

\title{XSGRB - The X-shooter GRB afterglow sample
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}}

@@

From March 2005 to March 2016 there have been about 350 {\it Swift} bursts
complying with our sample selection criteria (see ``Immediate objective''), and
about half of them have measured redshifts. Among the latter subset, the team
proposing these observations has measured about two thirds of the redshifts,
mainly with FORS1/2 and X-shooter \citep[see][]{Fynbo2009, Jakobsson2012, Kruhler2012}(see Fynbo et al.\ 2009, ApJS, 185, 526; 
Jakobsson et al.\ 2012, ApJ, 752, 62; Kr\"uhler et al.\ 2012, ApJ, 758, 46). Our
current aim is to build a sample superior to our previous low-resolution survey
(Fynbo et al.\ 2009, ApJS, 185, 526), both in terms of quantity and  quality of
the spectra. Our program started as guaranteed time observations during periods
84-91, and we have continued in open time since then.

@@

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}}
	\caption{Mollweide projection in galactic coordinates of the full sky showing
		the positions on the sky of the bursts presented in this work. The empty
		circles are the positions of all the \textit{Swift} bursts detected until the
		end of 2016. In blue is shown the position of the bursts fulfilling the sample
		criteria specified in Sect. \ref{samplecrit}. The red circles are added around
		GRBs which have been followed up by X-shooter since the commissioning of the
		instrument. Blue points with red circles represent GRBs that enter our sample
		and have X-shooter spectroscopy. The different samples are compared in Sect.
		\ref{results}. The background shows the dust maps presented in
		\citet{Schlegel1998} where the sample criteria cut with $A_V \lesssim 0.5$ mag
		is removed.}
	\label{fig:skymap}
\end{figure*}

@@

\begin{figure}[!ht]
	\centerline{\includegraphics[width=8cm]{figures/resolution_paper.eps}}
	\caption{Green datapoints show the FWHM (km/s) of Gaussian fits to unresolved telluric absorption lines in the VIS spectra, as a function of the FWHM of a Gaussian fit onto the trace in spatial direction at  792 nm. The lower horizontal axis is in units pixels, the top axis in arc seconds. The red datapoints show a subsample of NIR spectra.
		The grey line shows a linear fit to the VIS datapoints. }
	\label{fig:res}
\end{figure}

@@

Using the observational characteristics of the 1244 \todo{Get number} bursts
observed by \textit{Swift} to date \cite{Gehrels2009}, and the X-ray derived
hydrogen column densities \citep{Evans2009}, we can quantify the degree to which
our sample is biased against the overall \textit{Swift} sample in terms of
intrinsic GRB properties. The values are queried from the online \textit{Swift}
database maintained\footnote{http://swift.gsfc.nasa.gov/archive/grb\_table/}.
Three main samples are of interest to compare to assess the completeness of the
follow-up campaign; the full \textit{Swift} sample consisting of all the bursts
observed by \textit{Swift}, excluding the bursts entering our sample (1), the
burst that fulfill the selection criteria imposed in Sect. \ref{samplecrit}, but
has not been followed up (2), and the bursts actually followed up with X-shooter
(3).

@@

GRB follow-up requires the immediate availability of the telescope in order to
successfully carry out spectroscopic observations. Out of the 158 \todo{Get
exact number} bursts in the statistical sample, 32($\sim$ 20 percent) \todo{Get
exact number} have not been observed due to conditions, not relating the GRB or
afterglow properties. The reasons includes unavailability of the telescope due
to technical maintenance such af mirror re-coating, a visitor rejecting the TOO
trigger, or bad weather at the observing site. Because this cut is irrespective
of the GRB properties, it is not expected that this should change sample
properties. Addressing whether this cut introduces a bias in the followed sample
is not trivial, but in terms of intrinsic GRB properties, the sample
distributions does not change significantly, as shown in Sect. \ref{completeness}.
\todo{Should this be quantified somehow?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample darkness}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A subset of all GRBs exhibit no detectable or very faint optical afterglow
\citep{Groot1998, Djorgovski2001, Fynbo2001}, which had been parametrized in
term of "darkness" \citep{Jakobsson2004, VanderHorst2009}. The X-ray properties
of this subset has previously been investigated \citep{DePasquale2003,
	Melandri2012}, finding a slightly higher X-ray luminosity compared to the
optically brights bursts combined with similar prompt characteristics pointing
towards a denser circumburst material. This indicates along with investigations
of host galaxy extinction \citep{Greiner2011, Kruhler2011, Hjorth2012}, that the
extinction of the optical afterglow is primarily driven by the presence of dust,
local to the burst \todo{Local dust reference}. Because the dust content of a
galaxy is correlated with the gas amount \citep{Bohlin1978, Guver2009}, we
expect the X-ray derived column densities for the bursts without an optical
afterglow to be higher than the ones without.

For all bursts with follow-up within 100 hours we can calculate the
"darkness"-parameter, $\beta_{OX}$ \citep{Milvang-Jensen2012}. This required the
simultaneous measurement of the X-ray flux density and the optical flux density
which is in practice never possible. We use the measured acquisition camera
magnitude reported in Tab. \ref{tab:sample_overview} to get the optical flux
density at the beginning of the spectroscopic integration. Because we know the
delay between the follow-up and the \textit{Swift} trigger, we can use the
measured XRT burst lightcurve \citep{Evans2007,
	Evans2009}\footnote{http://www.swift.ac.uk/xrt\_curves/trigger\_numer/flux.qdp}
to infer the corresponding X-ray flux density. This is done by either linearly
interpolating between temporally neighboring XRT measurements of by
extrapolating the last few X-ray data points to the spectroscopic delay. When
the afterglow is not detected in the acquisition camera, an upper limit of $> 24$
mags is used, which propagates into an upper limit on $\beta_{OX}$.

In Fig. \ref{fig:betaOX} we compare the $\beta_{OX}$ - $N_H$ distribution with
the one presented in \citet{Fynbo2009}. We have replaced limits with values,
meaning that we artificially bias the distribution towards higher
$\beta_{OX}$-values because darker bursts naturally fainter in the optical
making the fraction of non-detections higher for lower $\beta_{OX}$. We
marginally confirm the result by \citet{Fynbo2009} that dark bursts $\beta_{OX}
< 0.5$) have slightly higher X-ray derived column densities, by calculating the
median, 14th, and 86th percentile of the distributions arising by separating the
sample at $\beta_{OX} = 0.5$. For bursts with $\beta_{OX} \geq 0.5$ we find
$\log(N_H) = 21.4_{-1.0}^{+0.7}$, whereas for $\beta_{OX} < 0.5$ we find
$21.8_{-0.9}^{+0.5}$, meaning that the median value for dark bursts is higher,
but with a great deal of overlap. A 2-sided KS test fails to reject the null
that they are drawn from the same distribution with $p = 0.43$, meaning there is
no clear evidence for a discrepancy.

\begin{figure}
	\centerline{\includegraphics[width=9cm]{figures/betaOX.pdf}}
	\caption{Optical extinction relative to X-ray against X-ray derived hydrogen
	column densities. In red is shown the sample presented in \citet{Fynbo2009}
	where is lines indicate the kernel density estimate of the distribution where
	limits have been replaced with values. Darker colors represented denser points.
	The correcspoing marginal distributions are shown. In blue is the points for
	the bursts presented here along with the marginal histograms. Limits on
	$\beta_{OX}$ is shown by downwards facing arrows.}
	\label{fig:betaOX}
\end{figure}

Because we only follow up targets with either a clear afterglow or counterpart,
correctly associating a galaxy with a burst when there is no detected optical afterglow
is more difficult, meaning that we should not have attempted follow-up for a
larger fraction of bursts with no optical afterglow. The X-ray properties for
the followed sample does not differ signicantly from the statistical sample, as
shown in Sect. \ref{completeness}. Using the table maintained by
Greiner\footnote{http://www.mpe.mpg.de/~jcg/grbgen.html} \todo{How to reference
this}, we can see how the pressence of an optical afterglow affects the
follow-up statistics. 50.5 percent of all \textit{Swift}-detected bursts do not
have an detected optical afterglow, but this number also includes bursts where
no optical telescopes were available, so is a lower limit. For the bursts
entering our statistical sample, this number is 27 percent

Since \textit{all} GRBs should have an optical afterglow, the presence of an optical afterglow is more a statement about the sensitivity than optical properties of the bursts.

@@

@@

This work made use of data supplied by the UK Swift Science Data Centre at the University of Leicester.
\end{acknowledgements}

@@

\clearpage

\begin{figure*}[!h]
	\centering
	\includegraphics[page=1, width=16cm]{figures/HI_measurements.pdf}
	\caption{Measurements of the hydrogen column-densities for all bursts with a
		clear Lyman alpha absorption system. In solid black is shown the spectrum with
		black dotted giving the corresponding 1-$\sigma$ error. Black dashed shows zero
		flux density. The solid red line is the absorption of column density equal to
		the value presented in Tab. \ref{tab:HI} with the 1-$\sigma$ interval shown
		with dashed lines.}
	\label{fig:HI1}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=2, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI2}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=3, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI3}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=4, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI4}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=5, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI5}
\end{figure*}
\clearpage

\appendix

@@

\subsection{GRB090313 (z = 3.373)}
The first GRB ever observed with X-shooter, during the commissioning of the
instrument, this data formed the basis of GCN
\#9015\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/9015.gcn3}} and is published
in \citet{DeUgartePostigo2010}. Due to the lingering brightness of GRB090313,
6.9 ks spectroscopic integration starting 45 hours after the BAT trigger reveals
a wealth of absorption features superposed on the afterglow continuum at a
common redshift of $z = 3.373$. Two intervening systems at $z = 1.959$ and $z =
1.800$ is identified based on strong \mgii-absorption. Because this burst is
observed before the instrument is science-verified, it does not enter into the
statistical sample.

\subsection{GRB090530 (z = 1.266)}
Observed during paranalization of the instrument, this data forms the basis of
GCN \#15571\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/15571.gcn3}}, but is not
published elsewhere. Observations began 20.6 hours after the BAT trigger and 4.8
ks spectroscopic integration in all three arms reveals the absorption signature
for a host at $z = 1.266$ from the detection of \mgii, \mgi, \SIii, \feii,
\aliii. Because this burst is observed before the instrument is
science-verified, it does not enter into the statistical sample.

\subsection{GRB090809 (z = 2.737)}
Observed during the first science verification period and was used for GCN
\#9761\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/9761.gcn3}} and is
additionally used as the basis for the master thesis by \'Asa Sk\'ulad\'ottir
(2010). 7.2 ks integration starting 10.2 hours after the GRB trigger notice
yields the clear afterglow continuum in all arms from with Lyman Limit located
in the beginning of the UVB coverage. The simultaneous detection of absorption
lines identified as \lya, \SIii, \oi, \SIi$^*$, \SIiv, \civ, \feii, \alii,
\aliii and \mgii at $z = 2.737$ sets it as the redshift of the GRB. Because this burst is observed before the instrument is
science-verified, it does not enter into the statistical sample.

\subsection{GRB090926 (z = 2.106)}
Starting during the second science verification period, this dataset forms the
basis of GCN \#9942\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/9942.gcn3}} and
is additionally published in \citet{DElia2010}. Spectroscopic integration
started 22 hours after the BAT trigger and from the acquistion camera the
optical afterglow has R = 17.9 mag (vega) at the beginning of the observations
which causes a very continuum to be seen in all arms. An absorption trough due
to \lya is clearly visible along with numerous metal resonance lines \civ,
\SIii, \SIii$^*$ \feii, \mgii, all at $z = 2.106$, marks this as the redshift of
the GRB. Because this burst is observed before the instrument is
science-verified, it does not enter into the statistical sample.

\subsection{GRB091018 (z = 0.971)}
The first burst observed during normal operation after science verification was
completed and there is the first burst that enter the statistical sample. This
data is the basis for GCN \#
10042\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10042.gcn3}} and is published
in \citet{Wiersema2012}. With a bright afterglow and a rapid follow-up, this
spectrum is of pristine quality. The afterglow continuum is bright throughout
all spectroscopic arms which allows the ready detection of \alii, \aliii, \feii,
\mnii, \mgii, \mgi, and \caii al located at $z = 0.971$, setting is as the
redshift of the host.

\subsection{GRB091127 (z = 0.490)}
Obtained 4 days after the burst trigger, this data forms the basis for GCN \#
10233\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10233.gcn3}} and is published
in \citet{Vergani2011}. Due to the late follow-up and a nearby moon, the S/N of
the afterglow continuum is poor especially in the UVB arm, why no clear
absorption lines are detected against the afterglow continuum, although see
\citet{Vergani2011} which report a tentative detection of \mgii. Emission lines
from the underlying host is clearly visible with lines from \oii, \hb, \oiii,
and \ha~all at $z = 0.490$. This bursts is additionally associated with
SN2009nz.

@@

\subsection{GRB100219A (z = 4.667)}
The data presented here also formed the basis of GCN \#
10441\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10441.gcn3}} and is published
in \citet{Thone2013}. Observations started 12.5 hours after the \textit{Swift}
trigger and has a total exposure time of 4.8 ks. Absorption features, including
those of \lya~and from a multitude of ions are detected against the afterglow
continuum at $z = 4.667$. Additionally, absorption from an intervening system
is found at $z = 2.181$.
